---
title: 'Project 3: Point patterns analysis'
author: 'Laura Sudupe. email: laura.medinilla@kaust.eu.sa'
date: "18/11/2022"
output:
  pdf_document: default
  html_document: default
fontsize: 12pt
geometry: margin=2cm
always_allow_html: yes

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include =FALSE}
library(spocc)
library(dplyr)
library(sp)
library(tmap)
library(raster)
library(rnaturalearth)
library(rmapshaper)
library(rgeos)
library(INLA)
library(ggplot2)
```


# Poland geospatial data analysis


## Abstract

**This project will be focused on predicting the intensity of the common kingfisher sighting in Spain. We are going to use the climate effect as a covariate. Point pattern  project is the thrird one of the three we need to prepare for the <a href = "https://www.paulamoraga.com/book-gds/index.html">STAT394 course</a>. In the first section we are going to describe the data and briefly explain point pattern data type. Next, we will focus on preparing the data for the statistical modelling. After this, we are going to go through the results, understanding the meaning and finally a short discussion of the project.**

```{r echo=F, warning= F, message=F}
#opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE, tidy = FALSE, cache = FALSE,results = 'asis' )
```


## Introduction

If we are dealing with data which has spatial continuity and have been collected at specific places we have geospatial data. In our case, this data represents the level of a pollutant, a particulate matter < 10 micrometers aerosol measured in different location in Poland during 2019. This particles are suspended in the air and has less than 10 micrometers of diameter. Some particles can get into the lungs and also in the bloodstream. The data had been downloaded from <a href = "http://aidef.apps.eea.europa.eu/?source=%7B%22query%22%3A%7B%22bool%22%3A%7B%22must%22%3A%5B%7B%22term%22%3A%7B%22Pollutant%22%3A%22Particulate%20matter%20%3C%2010%20%C2%B5m%20(aerosol)%22%7D%7D%2C%7B%22query_string%22%3A%7B%22query%22%3A%22Poland%22%2C%22default_operator%22%3A%22OR%22%2C%22analyze_wildcard%22%3Atrue%7D%7D%5D%7D%7D%2C%22display_type%22%3A%22tabular%22%7D">European Environment Agency (EEA)</a>.

The geospatial data of our pollutant is measured in a finite locations and we want to infer the characteristics of the spatial process such as the mean and variability of the process. This features are useful when predicting the process at no measured locations. We are going to model the data using *R-INLA*. 

We are going to use stochastic partial differential equation approach (SPDE) implemented in *R-INLA*. We are going to fit a spatial model and predict the variable of interest in no measure locations.


\newpage

## Data

Poland map was obtained from the **raster** package. We use **GADM**, the global administrative boundaries and level 0. In our case, we do not want to have more subdivisions in the country. 

We need to transform the object with **UTM** projection. For this, we use the ESPG code of Poland, 25830.


```{r data, echo = F, warning=FALSE}
aa <- occ(query = 'Alcedo atthis', from = 'gbif',
          date = c("2000-01-01", "2020-12-31"),
          gbifopts = list(country = "ES"),
          has_coords = TRUE, limit = 1000)
names(aa)
a <- occ2df(aa)
summary(a)

#We can visualize the locations of sloths retrieved in Spain using several mapping package

dpts_a <- SpatialPoints(a[, c("longitude", "latitude")])

#Then we create the map plotting the locations of dpts

tmap_mode("view")
tm_basemap() + tm_shape(dpts_a) + tm_dots()
```

The European Environment Agency provide air pollution measurements of the different European countries. For this proyect we use 2019 particulate matter (PM10) measurements. Data distribution on the 2019 year on Poland. After downloaded the data, we subset and filter it to obtain a normal distributed dataset to work with. 

We keep the variables indicating the day/year, monitoring station id, longitude, latitude and pollutant values.


```{r data_1, echo = F}
#4 Spatial climatic covariates

rmonth_a <- getData(name = "worldclim", var = "tmin", res = 10)
#plot(rmonth_a)

rcov_a <- mean(rmonth_a)
#plot(rcov_a)

#5 Implementing and fitting the spatial point process model
#5.1 Log-Gaussian Cox process model
#5.2 Computational grid

map_a <- ne_countries(type = "countries", country = "Spain", scale = "medium")

###take out canary islands
map_a <- ms_filter_islands(map_a, min_area = 12391399903)
#plot(map_a)

##############3

resolution_a <- 0.20
r_a <- raster(map_a, resolution = resolution_a)
print(r_a)

(nrow_a <- nrow(r_a))
(ncol_a <- ncol(r_a))
nrow_a*ncol_a

#set to 0 the values of all the grid cells
r_a[] <- 0
tab_a <- table(cellFromXY(r_a, dpts_a))
tab_a

r_a[as.numeric(names(tab_a))] <- tab_a
#plot(r_a)

#Finally, we convert the raster r to a SpatialPolygonsDataFrame
grid_a <- rasterToPolygons(r_a)
```

```{r figure_1,fig.align='center', echo = F, fig.cap="\\label{fig:figure_1}XXXXXXXX"}

plot(rcov_a)

```


```{r data_2, echo = F}
#5.3 Data
#Now, we include in the SpatialPolygonsDataFrame grid the data needed for modeling
grid_a <- grid_a[as.vector(t(matrix(1:nrow(grid_a), nrow = ncol_a, ncol = nrow_a))), ]

grid_a$id <- 1:nrow(grid_a)
grid_a$Y <- grid_a$layer
grid_a$cellarea <- resolution_a*resolution_a

grid_a$cov <- extract(rcov_a, coordinates(grid_a))

#Finally, we delete the cells of grid that fall outside Costa Rica
#plot(grid)
gridmap_a <- raster::intersect(grid_a, map_a)
grid_a <- grid_a[grid_a$id %in% gridmap_a$id, ]
#plot(grid_a)
summary(grid_a)
#missing data take out
indNA_a <- which(is.na(grid_a$cov))
indNA_a
grid_a$cov[indNA_a] <- grid_a$cov[indNA_a+1]

#We use tmap to create maps of the number of sloths and 
#the covariate values in each of the cells. 

gridborder_a <- gUnaryUnion(grid_a)

```




```{r figure_2,fig.align='center', echo = F, fig.cap="\\label{fig:figure_2}XXXXXX"}

tmap_mode("plot")
tm_shape(grid_a) +
  tm_polygons(col = c("Y", "cov"), border.col = "transparent") +
  tm_shape(gridborder_a) + tm_borders() +
  tm_facets(ncol = 2) + tm_legend(legend.position = c("left", "bottom"))

```





\newpage

## Modelling

### Mesh contruction



\newpage

## Model

We are going to specify a spatial model to predict PM10 and fit the model using **R-INLA**. PM10 measurements are at different stations in Poland. yet, we have continuity in space, so we can use a geostatistical model to predict PM10 values in other points in the map.

We assume the PM10 meditions $Y_{i}$ at $s_{i}$ locations follows a normal distribution with mean $\mu_{i}$ and variance $\sigma^{2}$. The mean $\mu_{i}$ is expressed as the sum of and intercept $\beta_{0}$ and a spatially structured random effect that follows a zero-mean Gaussian process.




$$Yit \sim N(\mu_{it},\sigma^{2}_{e} i) $$


$$\mu_{it} = \beta_{0} + \xi( x_{i},t)$$


Where, 

- $Yit$ the PM10 values measured at location i,
- $B0$ is the intercept,
- $\sigma^{2}_{e}$ is the variance of the measurement error,
- $\xi( x_{i},t)$ is a spatial random effect at location $x_{i}$.


The response variable is in the left hand side of the formula, the fixed and random effects on the right side. Last step is to fit the model by inla(), using the default priors.


```{r formula, echo=F, message=FALSE}
#5.4 Fitting the model using INLA

grid_a$id2 <- grid_a$id

formula <- Y ~ 1 + cov +
  f(id, model="rw2d", nrow = nrow_a, ncol = ncol_a) +
  f(id2, model="iid")

res_a <- inla(formula, family = "poisson", data = grid_a@data,
            E = cellarea_a, control.predictor = list(compute = TRUE))


```


\newpage

## Results and Interpretation

With summary(res) we can see the parameter estimates of the fixed and random effects. The intercept has *21.78* mean value and *0.375* standard deviation with 95% credible interval between -20.957 and 22.442.

The data frame res$summary.fitted.values contains the mean, and the 2.5 and 97.5 percentiles of the fitted values. The indices of the rows corresponding to the predictions can be obtained with the inla.stack.index() function by passing stk.full and specifying the tag "pred".

```{r results, echo=F}
#5.5 Results
summary(res_a)

marginal_a <- inla.smarginal(res_a$marginals.fixed$cov)
marginal_a <- data.frame(marginal_a)
ggplot(marginal_a, aes(x = x, y = y)) + geom_line() +
  labs(x = expression(beta[1]), y = "Density") +
  geom_vline(xintercept = 0, col = "black") + theme_bw()
nrow(res_a$summary.random$id)
grid_a$respa <- res_a$summary.random$id[grid_a$id, "mean"]

#We can also obtain the posterior mean of the unstructured random effect.
grid_a$reiid <- res_a$summary.random$id2[, "mean"]

```




```{r figure_3,fig.align='center', echo = F, fig.cap="\\label{fig:figure_3}XXXXXXX"}

#Then we can create maps of the random effects with tmap
tm_shape(grid_a) +
  tm_polygons(col = c("respa", "reiid"), style = "cont", border.col = "transparent")  +
  tm_shape(gridborder_a) + tm_borders() +
  tm_facets(ncol = 2) + tm_legend(legend.position = c("left", "bottom"))


```




```{r figure_4,fig.align='center', echo = F, fig.cap="\\label{fig:figure_4}XXXXXXX"}

cellarea_a <- resolution_a*resolution_a
grid_a$NE <- res_a$summary.fitted.values[, "mean"] * cellarea_a
grid_a$LL <- res_a$summary.fitted.values[, "0.025quant"] * cellarea_a
grid_a$UL <- res_a$summary.fitted.values[, "0.975quant"] * cellarea_a
summary(grid_a)

#We use tmap to create maps with the mean and lower and upper limits of 95% credible

tm_shape(grid_a) +
  tm_polygons(col = c("NE", "LL", "UL"),
              style = 'fixed', border.col = "transparent",
              breaks = c(0, 1,2, ceiling(max(grid_a$UL)))) +
  tm_shape(gridborder_a) + tm_borders() +
  tm_facets(ncol = 3) + tm_legend(legend.position = c("left", "bottom"))
```




We can see in Figure 6 the posterior distributions of the intercept.


We plot the predicted PM10 values for 2019 year. To be able to obtain the predictions from the model, we use inla.stack.index() function to extract the indices. We use the mean predictions with the lower and upper limits of the 95% credible intervals and provide measures of uncertainty. 


If we observe the intensity changes overall, the spatial measurements gradients are similar. If we focus on the 95% limits, lower limit map is showing less pollutant levels than the upper limit. 

The posterior mean for the predicted PM10 values in 2019 is lower in the north. The spatial effect is higher in the south, meaning in higher posterior mean in the south. The correlation with the polutant and the ocean proximity could be deeper analyze in next studies.


\newpage

### Discussion section


The maps give us information about the spatial predictions the year the the data was collected, the limitation of the study is the temporal absence, the data had been measured more years but for this study we only use 2019 data. To improve this study we can produced a spatio-temporal model. This maps can help us to understand the cities pollutant concentrations if we mixed this information with other variables, as cities location or population.

The proximity to the ocean is also a variable to take into account

With this project I learned how to download, manipulate and analyze geospatial data. I work with different packages to do so, gaining experience with Rstudio, Rmarkdown and R-INLA for example.

  
  
  

