---
title: 'Project 3: Point patterns analysis'
author: 'Laura Sudupe. email: laura.medinilla@kaust.eu.sa'
date: "18/11/2022"
output:
  pdf_document: default
  html_document: default
fontsize: 12pt
geometry: margin=2cm
always_allow_html: yes

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include =FALSE}
library(spocc)
library(dplyr)
library(sp)
library(tmap)
library(raster)
library(rnaturalearth)
library(rmapshaper)
library(rgeos)
library(INLA)
library(ggplot2)
```


# Poland geospatial data analysis


## Abstract

**This project will be focused on predicting the intensity of the common kingfisher sighting in Spain. We are going to use the climate effect as a covariate. Point pattern  project is the third one of the three we need to prepare for the <a href = "https://www.paulamoraga.com/book-gds/index.html">STAT394 course</a>. In the first section we are going to describe the data and briefly explain point pattern data type. Next, we will focus on preparing the data for the statistical modelling. After this, we are going to go through the results, understanding the meaning and finally a short discussion of the project.**

```{r echo=F, warning= F, message=F}
#opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE, tidy = FALSE, cache = FALSE,results = 'asis' )
```


## Introduction

For stochastic models we use a set of random variables, projections and calculations are repeated to reach the result probability distributions. In our case, we have point located in different places in the maps that give us information about a event. What we can achieve is the identification of pattern in the distribution and evaluate the intensity. When working with spatial point patterns we also want to study the correlation between the points location and spatial covariates.

In our case, the data represents common kingfisher sighting in Spain. Also call the river kingfisher, is distributed across Eurasia and North Africa. The data had been downloaded from the package *spocc*, where we have the location of the sighting between 2000-2020. In this project, We will work with a fit log-Gaussian Cox process model where we will use the anual minimum temperature in Spain as a covariate and random effect to model the unexplained variability. We are going to model the data using *R-INLA*. 


\newpage

## Data

Kingfisher data was obtained from the **spocc** package.  We specify country = ES to select Spain occurrences and select has_coords = TRUE in order to get the coordinates. We get a *occdat* class object to work with.


```{r data, echo = F, warning=FALSE}
aa <- occ(query = 'Alcedo atthis', from = 'gbif',
          date = c("2000-01-01", "2020-12-31"),
          gbifopts = list(country = "ES"),
          has_coords = TRUE, limit = 1000)
#names(aa)
a <- occ2df(aa)
#summary(a)

#We can visualize the locations of sloths retrieved in Spain using several mapping package
dpts_a <- SpatialPoints(a[, c("longitude", "latitude")])

```


The interactive map show us the kingfisher locations. There is a heterogeneous distribution with more point near the sea. This could be due to high abundance of rivers near the sea. Even so, there are regions with more points like near the Pirineus and Extremadura regions.

```{r figure_1,fig.align='center', echo = F,warning=FALSE,fig.cap="\\label{fig:figure_1}Kingfisher ocurrence in Spain"}
#Then we create the map plotting the locations of dpts

tmap_mode("view")
tm_basemap() + tm_shape(dpts_a) + tm_dots()

```


## Spatial climatic covariates


We are going to use the annual minimum temperature observed in Spain. Kingfishers used to migrate to different areas if the river where they live get frozen. So this spatial variable can influence the occurrence. The temperature information was obtained from *raster* package. 

We get a object with minimum temperature for each month. So we transform the data to obtain the annual average minimum temperature.


```{r data_1, echo = F, warning=FALSE}
#Spatial climatic covariates

rmonth_a <- getData(name = "worldclim", var = "tmin", res = 10)
#plot(rmonth_a)

rcov_a <- mean(rmonth_a)
#plot(rcov_a)

#5 Implementing and fitting the spatial point process model
#5.1 Log-Gaussian Cox process model
#5.2 Computational grid

map_a <- ne_countries(type = "countries", country = "Spain", scale = "medium")

###take out canary islands
map_a <- ms_filter_islands(map_a, min_area = 12391399903)
#plot(map_a)

##############3

resolution_a <- 0.35
r_a <- raster(map_a, resolution = resolution_a)
#print(r_a)

nrow_a <- nrow(r_a)
ncol_a <- ncol(r_a)
#nrow_a*ncol_a

#set to 0 the values of all the grid cells
r_a[] <- 0
tab_a <- table(cellFromXY(r_a, dpts_a))
#tab_a

r_a[as.numeric(names(tab_a))] <- tab_a
#plot(r_a)

#Finally, we convert the raster r to a SpatialPolygonsDataFrame
grid_a <- rasterToPolygons(r_a)
```



```{r figure_2,fig.align='center', echo = F, warning= F, fig.cap="\\label{fig:figure_2}Average anual min temperature"}

plot(rcov_a)

```


\newpage

# Implementing and fitting the spatial point process model

## Log-Gaussian Cox process model


We want to see patterns in the distribution of the sighting and estimate the intensity of events. Even if we can use Poisson process model for it, this model is too restrictive. So we are going to use log-Gaussian Cox process model, used in environmentally affected conditions.

We are assuming that the sighting locations in Spain, ${s_{i}:i=1,…,n}$ , has been generated as a log-Gaussian Cox process with intensity given by $\Lambda(s)=exp(\eta(s))$. We can use a grid approach like in the second project of this course, the n model can be fitted by latent Gaussian model through the gridding technique.

We want to discretize the region of interest into grid with $n_{1}×n_{2}=N$ cells 
${s_{ij}, i=1,…,n_{1},, j=1,…,n_{2}}$. The mean number of events in cell $s_{ij}$ is given by the integral of the intensity over the cell $\Lambda(ij)=\int_{s_{ij}}exp(\eta(s))ds$, and this integral can be approximated by $\Lambda(ij)\sim|s_{ij}|exp(\eta_{ij})$, where $|s_{ij}|$ is the area of the cell $s_{ij}$. Then, conditional on the latent field $\eta_{ij}$, the observed number of locations in grid cell $s_{ij}$, $y_{ij}$, are independent and Poisson distributed as follows:



$$Y_{ij}|\eta_{ij} \sim Poisson(|s_{ij}|exp(\eta_{ij})) $$


$$\eta_{ij} = \beta_{0} + \beta_{1} x  cov(s_{ij}) + f_{s}(s_{ij}) + f_{u}(s_{ij}) $$


Where, 

- $Yit$ the PM10 values measured at location i,
- $B0$ is the intercept,
- $cov(s_{ij})$ is the covariate at $s_{ij}$.
- $\beta_{1}$ is a the coefficient of $cov(s_{ij})$.
- $f_{s}$ is a spatially structured random effect reflecting unexplained variability.
- $f_{s}$ is a unstructured random effect reflecting independent variability in cell $s_{ij}$. 


## Computational grid and data 

We want to fit and project the model in the grid, so we create a grid that covers the main area of Spain. With ne_countries() function of *rnaturalearth* package we obtain the map and with ms_filter_islands() function we select only the main area, with out the islands. After we create a raster that covers Spain region. Finally, we convert the object to *SpatialPolygonsDataFrame*, we are going to use this grid to fit the model with INLA.

Once this is done, we nned to include in the grid the data we are going to use for modelling. We are going to add the following variables:

- *id* with the id of the cells.
- *Y* with the number of sloths.
- *cellarea* with the cell areas.
- *cov* with the covariate values in each of the cells.

For the missing values in the covariate, we input the covariate mean values. With *tmap* we create the plots of the number of kingfishers and the covariate values in each of the grid cells.  



```{r data_2, echo = F, warning=FALSE}
#Now, we include in the SpatialPolygonsDataFrame grid the data needed for modeling
grid_a <- grid_a[as.vector(t(matrix(1:nrow(grid_a), nrow = ncol_a, ncol = nrow_a))), ]

grid_a$id <- 1:nrow(grid_a)
grid_a$Y <- grid_a$layer
grid_a$cellarea <- resolution_a*resolution_a

grid_a$cov <- raster::extract(rcov_a, coordinates(grid_a))

#Finally, we delete the cells of grid that fall outside Spain
#plot(grid)
gridmap_a <- raster::intersect(grid_a, map_a)
grid_a <- grid_a[grid_a$id %in% gridmap_a$id, ]
#plot(grid_a)
#summary(grid_a)
#missing data take out
indNA_a <- which(is.na(grid_a$cov))
#indNA_a

grid_a$cov[indNA_a] <- grid_a$cov[mean(indNA_a)]

#We use tmap to create maps of the number of sloths and 
#the covariate values in each of the cells. 

gridborder_a <- gUnaryUnion(grid_a)

```




```{r figure_3,fig.align='center', echo = F,warning=FALSE, message=FALSE, fig.cap="\\label{fig:figure_3}Number of kingfishers and the covariate values"}

tmap_mode("plot")
tm_shape(grid_a) +
  tm_polygons(col = c("Y", "cov"), border.col = "transparent") +
  tm_shape(gridborder_a) + tm_borders() +
  tm_facets(ncol = 2) + tm_legend(legend.position = c("left", "bottom"))

```

\newpage

## Fitting the model with INLA

We are going to use *R-INLA* to fit the log-Gaussian Cox process to model the sighting data. First we need to specify the formula with the linear predictor and then call inla() function. 

*Y* is the outcome variable (number of occurrences in each cell of the grid). *f()* function is for the random effects. *id* is for the spatially structured effect and *id2* for the unstructured random effect.


The outcome variable is Y (the number of occurrences in each cell) and the covariate is cov.

The random effects are specified with the f() function where the first argument is an index vector specifying which elements of the random effect apply to each observation, and the other arguments are the model name and other options. In the formula, different random effects need to have different indices vectors. We use grid$id for the spatially structured effect, and create an index vector grid$id2 with the same values as grid$id for the unstructured random effect.

```{r formula, echo=T, message=FALSE}
#random effects
grid_a$id2 <- grid_a$id

#formula
formula <- Y ~ 1 + cov +
  f(id, model="rw2d", nrow = nrow_a, ncol = ncol_a) +
  f(id2, model="iid")
#fit the model
res_a <- inla(formula, family = "poisson", data = grid_a@data,
            E = cellarea_a, control.predictor = list(compute = TRUE))

```



## Results and Interpretation

With summary(res_a) we can see the parameter estimates of the fixed and random effects. The intercept  $B_{0}$ has *-4.892* mean value and *0.754* standard deviation with 95% credible interval between -6.409 and -3.444.

The data frame res_a$summary.fitted.values contains the mean, and the 2.5 and 97.5 percentiles of the fitted values. We can appreciate the positive effect of the covariate minimum temperature on the intensity of the process with a posterior mean  of 0.047 with a 95% credible interval between  0.034 and 0.060.

```{r results, echo=F, warning=FALSE}
head(summary(res_a))

marginal_a <- inla.smarginal(res_a$marginals.fixed$cov)
marginal_a <- data.frame(marginal_a)
#We can also obtain the posterior mean of the unstructured random effect.
grid_a$reiid <- res_a$summary.random$id2[, "mean"]
```

\newpage

We can see the posterior distribution of the coefficient of the covariate $B_{1}$ in the Figure 4.

```{r figure_4,fig.align='center', echo = F, warning=FALSE, message=FALSE, fig.cap="\\label{fig:figure_4}Posterior distribution of the coefficient of the covariate"}
ggplot(marginal_a, aes(x = x, y = y)) + geom_line() +
  labs(x = expression(beta[1]), y = "Density") +
  geom_vline(xintercept = 0, col = "black") + theme_bw()
nrow(res_a$summary.random$id)
grid_a$respa <- res_a$summary.random$id[grid_a$id, "mean"]
```

\newpage


We can can add to our grid the information of the posterior mean of the spatial effect in the cells of Spain in the Figure 5.

The pattern of the spatial structured random effect is not very constant, The unstructured random show us different locations with high values. 



```{r figure_5,fig.align='center', echo = F, warning=FALSE, message=FALSE, fig.cap="\\label{fig:figure_5}Grid the information of the posterior mean of the spatial effect"}

#Then we can create maps of the random effects with tmap
tm_shape(grid_a) +
  tm_polygons(col = c("respa", "reiid"), style = "cont", border.col = "transparent")  +
  tm_shape(gridborder_a) + tm_borders() +
  tm_facets(ncol = 2) + tm_legend(legend.position = c("left", "bottom"))


```




Figure 6 show us the mean and the quantiles of the predicted intensity (mean number of events per cell in the grid).  We use the mean predictions with the lower and upper limits of the 95% credible intervals and provide measures of uncertainty. 


We create a variable with the mean number of events of each cell, *NE*, where we store the predicted intensity multiplied by the cell area and added to our grid object. We also added the lower and upper limit credible intervals (95%) by adding the quantile values multiplied by the cell areas.


```{r figure_6,fig.align='center', echo = F, warning=FALSE, message=FALSE, fig.cap="\\label{fig:figure_6}Intersity of kingfisher ocurrence"}

cellarea_a <- resolution_a*resolution_a
grid_a$NE <- res_a$summary.fitted.values[, "mean"] * cellarea_a
grid_a$LL <- res_a$summary.fitted.values[, "0.025quant"] * cellarea_a
grid_a$UL <- res_a$summary.fitted.values[, "0.975quant"] * cellarea_a
#summary(grid_a)
#We use tmap to create maps with the mean and lower and upper limits of 95% credible
tm_shape(grid_a) +
  tm_polygons(col = c("NE", "LL", "UL"),
              style = 'fixed', border.col = "transparent",
              breaks = c(0, 3,6,9, ceiling(max(grid_a$UL)))) +
  tm_shape(gridborder_a) + tm_borders() +
  tm_facets(ncol = 3) + tm_legend(legend.position = c("left", "bottom"))


```


If we observe the intensity changes overall, most of our cells has low values, the highest values are 12 kingfishers in the cell. The maps with the lower and upper limits of 95% credible intervals denote the uncertainty of these predictions.        


\newpage

### Discussion section

Those plots give us information about the intensity predictions of the years the data was collected. 

The kingfisher lives in the rives, the limitation of the study is the river spatial location absence.  To improve this study we can produced model where we study the river occurrence. This maps can help us to understand this specie habitat when we mixed this information with other variables, as river location or temperature.

With this project I learned how to download, manipulate and analyze point patterns data. I work with different packages to do so, gaining experience with Rstudio, Rmarkdown and R-INLA for example.

  
  
  

