---
title: 'Project 1: Areal data analysis'
author: 'Laura Sudupe. email: laura.medinilla@kaust.eu.sa'
date: "10/1/2022"
output:
  pdf_document: default
  html_document: default
fontsize: 12pt
geometry: margin=2cm
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include =FALSE}
library(sf)
library(epitools)
library(SpatialEpi)
library(spData)
library(ggplot2)
library(leaflet)
library(spdep)
library(INLA)
library(tidyverse)
library(palmerpenguins)
library(ggdist)
library(gridExtra)
library(PerformanceAnalytics)
```


# New Zeland aeral data analysis


### Abstract

**This project will be focused on analyzing New Zeland (NZ) aeral data using Rstudio. The project is the first one of the three we need to prepare for the <a href = "https://www.paulamoraga.com/book-gds/index.html">STAT394 course</a>. In the first section we are going to describe NZ data and briefly explain areal data type. Next, we will focus on preparing the data for the statistical modelling. After this, we are going to go through the results, understanding the meaning and finally a short discussion of the project.**

```{r echo=F, warning= F, message=F}
#opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE, tidy = FALSE, cache = FALSE,results = 'asis' )
```


### Introduction

When we are working with a fixed domain which is divided in a finite number or sub regions where we have aggregated data, we can say we are working with Aeral data. This type pf data is presented as attributes of fixed polygonal areas, this areas doesn't need to be regular shaped. We can apply statistical methods to extract attributes and analyze then in space.


As part of the STAT394 course, in this project we are going to focus on nz (spData), where we have regions in New Zealand and as part of the course, we are going to model the data using INLA.


### Data

Brief description of the dataset, we can see it is a sf object

```{r data, include=FALSE}
data("nz")
head(nz)
str(nz)
?nz

#https://www.ehinz.ac.nz/indicators/uv-exposure/melanoma/
cases <- read.csv("./new-cancer-registrations-2017-data-files-dec19/cancer.csv", sep = ",")
nz$cases <- cases$cases
```

```{r statistical_description, echo = F}
class(nz)
names(nz)
```

The dataset is "sf" class and it contains 16 observations and 7 variables. The followings:

*- Name: The county name*

*- Island: North or Shout location of the county*

*- Land_area: The area of each county*

*- Population: The total population of each county*

*- Median_income: Median income of the county*

*- Sex_ratio: Ratio in each of the counties*

*- Geom: The variable with the spatial information*


In our case, we are going to focus on the cancer cases and try to see the dependence of it with the sex_Ratio in each county. In the next plots we can observe how are our variables distributed. For the population, we can see almost all the points are bellow 500 000. We have a pick in the South with more counties with low population, the north counties are more homogeneous. If we focus on the Median_income, the South shows most of the point between 25 000 and 30 000. In the north, the range of values is larger. Even though, the south counties show us more homogeneous distribution of the Land_area than the North. Related to the cases, the south is more heterogeneously distributed than the north.

```{r statistical_description_1, echo=F}
a <- nz %>%
  ggplot(aes(x=Population, y=Island, fill=Island))+
  stat_slab(aes(thickness = stat(pdf*n)), 
                scale = 0.3) +
  stat_dotsinterval(side = "bottom", 
                    scale = 0.3, 
                    slab_size = NA) 
b <- nz %>%
  ggplot(aes(x=Median_income, y=Island, fill=Island))+
  stat_slab(aes(thickness = stat(pdf*n)), 
                scale = 0.3) +
  stat_dotsinterval(side = "bottom", 
                    scale = 0.3, 
                    slab_size = NA)
c <- nz %>%
  ggplot(aes(x=cases, y=Island, fill=Island))+
  stat_slab(aes(thickness = stat(pdf*n)), 
                scale = 0.3) +
  stat_dotsinterval(side = "bottom", 
                    scale = 0.3, 
                    slab_size = NA) 

d <- nz %>%
  ggplot(aes(x=Sex_ratio, y=Island, fill=Island))+
  stat_slab(aes(thickness = stat(pdf*n)), 
                scale = 0.3) +
  stat_dotsinterval(side = "bottom", 
                    scale = 0.3, 
                    slab_size = NA) 

grid.arrange(a, b, c, d,  ncol=2)
```


In this data we have the sex_ratio. This variable give us the ratio of female to males in a population, it tends to be 1:1. In our case, we are going to transform the data

*-Sex_ratio = 1, then "same"*

*-Sex_ratio > 1, then "female"*

*-Sex_ratio < 1, then "male"*


```{r statistical_description_3, echo=F, message=FALSE, warning=FALSE}
nz$Island <- as.factor(nz$Island)
##https://rdrr.io/cran/SpatialEpi/src/R/expected.R
nz$sex <- ifelse(nz$Sex_ratio == 1, 'same', ifelse(nz$Sex_ratio > 1, 'female', "male"))
nz$Population <- as.integer(nz$Population)

###expected cases
##area: name of the area
##Y: observed number of cases, in my case, median income
##E: expected number of cases in my case, expected median income
##I want it to depend on
##population in each 
##SMIR in each area

```


In the next plot we check the correlation between the principal variables. There is a correlation between the Land_area and the Sex_ratio. When Land_area values is decreasing, the Sex_ratio is also decreasing, positive correlation. Overall, we are not seeing a bit correlation values between variables. In this work we are going to focus on the Median_income as a response variable, we can see a 0.44 correlation score between the Median_income and Population variables.


```{r statistical_description_2, echo=F, message=FALSE, warning=FALSE}
nz$Population <- as.numeric(nz$Population)
nz$Median_income <- as.numeric(nz$Median_income)
nz$Sex_ratio <- as.numeric(nz$Sex_ratio)
nz$Land_area <- as.numeric(nz$Land_area)
nz$cases <- as.numeric(nz$cases)
b<- nz[,3:6] 
b$cases <- nz$cases
b$geom <- NULL
chart.Correlation(b, histogram = TRUE, method = "pearson")
```



Along with the observed values, we want to know the expected ones. For this, we use the expected() function of the SpatialEpi package. For this analysis, we want to calculate the expected median_income. We divided the data between sex and hemisphere

```{r statistical_description_4, echo=T}
#Expected 
nz$e <- expected(population = nz$Population, cases = nz$cases, n.strata = 4)
```

```{r statistical_description_5, echo=F}
nz$SMR <- nz$cases / nz$e
```


Let's see our final variables in a spatial plot. We can see that only in one of the counties the ratio between female and males is higher for females


```{r summary, message=FALSE, echo=F}
if (requireNamespace("sf", quietly = TRUE)) {
  library(sf)
  summary(nz)
  plot(nz)
}
```


In our case, the map does not have the Coordinate Reference System (CRS) data. For this reason, we need to add it by ourself. Also, we need to convert the data to SpatialPolygonsDataFrame


```{r spatial, message=FALSE, echo=T}
#convert to SpatialPolygonsDataFrame
map_spdf <- as_Spatial(nz, IDs = TRUE) 
wgs84 = '+proj=longlat +datum=WGS84'
nz_region <- spTransform(map_spdf, CRS(wgs84))

class(map_spdf)
```


### Visual plots of the data

Let's focus on the SMR, We can see Auckland and Bay of plenty counties in the North are the ones with highest values. In the south, Otago has high value of SMR also.


```{r SMR, message=FALSE, echo=F}
###Mapping SMR
le <- leaflet(nz_region) %>% addTiles()
pale <- colorNumeric(palette = "YlOrRd", domain = nz_region$SMR)

le %>% addPolygons(color = "grey", weight = 1, fillColor = ~pale(SMR), fillOpacity = 0.5) %>%
  leaflet::addLegend(pal = pale, values = ~SMR, opacity = 0.5, title = "SMR", position = "bottomright")
```

Related to the sex ratio, female/male relation. It is obvious in the south we have higher ratio than in the north. The west coast has the highest value compared with the other counties. Is also the only county where the ratio is higher than 1.

```{r Sex_ratio, message=FALSE, echo=F}
###Mapping median income
le <- leaflet(nz_region) %>% addTiles()
  pale <- colorNumeric(palette = "YlOrRd", domain = nz_region$Sex_ratio)

le %>% addPolygons(color = "grey", weight = 1, fillColor = ~pale(Sex_ratio), fillOpacity = 0.5) %>%
  leaflet::addLegend(pal = pale, values = ~Sex_ratio, opacity = 0.5, title = "Sex_ratio", position = "bottomright")
```



### Analyses

For this proyect, we are going to use <a href = "https://www.r-inla.org/">INLA package</a> to model the data. This package is used to do approximate Bayesian inference for Latent Gaussian Models

In our case, $Yi$ is going to be the observed cases by county and $Ei$ the expected cases. $θi$ is the relative risk for county i=1,…,n. We are going to use Poisson model:

$$Yi|θi \sim Poisson(Eixθi), i=1, ... , n, $$

$$log(θi)=β0+β1×SexRatioi+ui+vi $$

- $β0$ is the intercept,
- $β1$ is the coefficient of the sex_ratio  covariate,
- $ui$ is a structured spatial effect,
- $vi$ is an unstructured spatial effect, vi∼N(0,1/τv).


We need to define some parameters before running the model. First, we need to created a neighborhood matrix where we define the spatial random effect. With poly2nb() function we create a neighbors list based on the areas surrounding the boundaries. For each area, we have the indices of its neighbors. After this, with nb2INLA() we transform the list in a file representing the matrix. R-INLA package need this file to work with to be able to specify the spatial random effect.


```{r modelling_1, message=FALSE, echo=F, include=FALSE}
nb_nz <- poly2nb(nz_region)
head(nb_nz)
nb_nz[[1]] #the first county has 2 neighbours

nb2INLA("nz_region.adj", nb_nz)
ge <- inla.read.graph(filename = "nz_region.adj")
```



For the model, we need to create two variables, where we are going to have the random effect values. $ui$ to model the spatial residual variation, and $vi$ to model unstructured noise.



```{r modelling_2, message=FALSE, echo=F, include=FALSE}
###inference using inla
#create variables
nz_region$re_u <- 1:nrow(nz_region@data)
nz_region$re_v <- 1:nrow(nz_region@data)
```


Define the formula

```{r modelling_3, message=FALSE, echo=T}
formula <- cases ~ Sex_ratio + 
  f(re_u, model = "besag", graph = ge, scale.model = TRUE) + 
  f(re_v, model = "iid")
```


Let's fit the model

```{r modelling_4, message=FALSE, echo=T}
#https://groups.google.com/g/r-inla-discussion-group/c/EPTiPRE7jAM?pli=1 ERROR
rese <- inla(formula, family = "poisson", data = nz_region@data,
             E = e,
             control.predictor = list(compute = TRUE),
             control.compute=list(return.marginals.predictor=TRUE))

```



### Results and Interpretation

We can focus on the intercept  $β0$= 0.952 value with 95% credible interval between -22.460 and 24.263. The coeficient for Population is $β̂ 1$ -2.4.

```{r results, message=FALSE, echo=F}
head(summary(rese))
```



```{r results_1, message=FALSE, echo=F}
library(faraway)
library(gridExtra)
#http://julianfaraway.github.io/brinla/examples/chicago.html
#https://github.com/julianfaraway/brinla
#remotes::install_github("julianfaraway/brinla")
library(brinla)
bri.hyperpar.summary(rese)
bri.hyperpar.plot(rese)
bri.fixed.plot(rese)
```

Let's see the posterior distribution of the sex_ratio distribution. We do this by calculating a smoothing of the marginal distribution of the coefficient with inla.smarginal() and then plot it with ggplot() of the ggplot2 package.

In this plot we can see if the parameter is different to 0. For the sex_ratio, we can see the posterior density is away from zero

```{r results_2, message=FALSE, echo=F}
###results
rese$summary.fitted.values

marginale <- inla.smarginal(rese$marginals.fixed$Sex_ratio)
marginale <- data.frame(marginale)
ggplot(marginale, aes(x = x, y = y)) + geom_line() + labs(x = expression(beta[1]), y = "Density") +
  geom_vline(xintercept = 0, col = "blue") + theme_bw()
```

We can get the risk estimation and the uncertainty for each of the areas. This information is given to us like the mean posterior and the 95% credible intervals of $θi$, i=1,…,n which are in res$summary.fitted.values.

So, the mean is the estimation of the relative risk, and the quantiles are the lower and upper limits of the 95% credible intervals of the risks. If we plot this results, we can see Auckland and Bay of plenty counties has higher relative risk values than the other counties in the north. From the south, Otago has the highest relative risk.

```{r results_3, message=FALSE, echo=F}
###add results to map
#head(rese$summary.fitted.values)

nz_region$RR <- rese$summary.fitted.values[, "mean"]
nz_region$LL <- rese$summary.fitted.values[, "0.025quant"]
nz_region$UL <- rese$summary.fitted.values[, "0.975quant"]

###mappring median income
pale <- colorNumeric(palette = "YlOrRd", domain = nz_region$RR)

labels <- sprintf("<strong> %s </strong> <br/> Observed: %s <br/> Expected: %s <br/>
                  Sex ratio: %s <br/>SMR: %s <br/>RR: %s (%s, %s)",
                  nz_region$Name, nz_region$Median_income,  round(nz_region$e, 2),  
                  nz_region$Sex_ratio, round(nz_region$SMR, 2),
                  round(nz_region$RR, 2), round(nz_region$LL, 2), round(nz_region$UL, 2)) %>%
  lapply(htmltools::HTML)

leaflet(nz_region) %>% addTiles() %>%
  addPolygons(color = "grey", weight = 1, fillColor = ~pale(RR),  fillOpacity = 0.5,
              highlightOptions = highlightOptions(weight = 4),
              label = labels,
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "15px", direction = "auto")) %>%
  leaflet::addLegend(pal = pale, values = ~RR, opacity = 0.5, title = "RR", position = "bottomright")
```




### Discussion section


What can I do better

  - More variables to check
  - Low number of data points to proper work
  
  
**With this project I learned how to download, manipulate and analize areal data. I work with different packages to do so,  **
What I learned

  - Manipulate spatial data
  - Show results in spatial data
  - Use INLA package
  - Deliver results
  
General comment

  - 
  
### Doubts

- am I defining the prior?
- is this enough?
- point me to better explain the results

  
  

