---
title: 'Project 1: Areal data analysis'
author: "Laura Sudupe. email: laura.medinilla@kaust.eu.sa"
date: "10/1/2022"
geometry: margin=2cm
fontsize: 12pt
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# New Zeland aeral data analysis


### Abstract

**I claim absolutely no credit for this visualization, which I consider one of the most best I have ever seen.  All credit belongs to the <a href = "http://www.nytimes.com/interactive/2011/05/31/business/economy/case-shiller-index.html">original source</a>.  If anybody believes this to be not fair use, I will take it down immediately.  I am implicitly assuming approval for this fork due to the <a href = "http://datastori.es/data-stories-22-nyt-graphics-and-d3-with-mike-bostock-and-shan-carter/">data.stories interview</a>.** 

```{r echo=F, warning= F, message=F}
#opts_chunk$set(message = FALSE, warning = FALSE, error = FALSE, tidy = FALSE, cache = FALSE,results = 'asis' )
```


### Introduction


### Data



### Analyses


### Results and Interpretation



### Discussion section

